---
title: 'Assignment 3: Logistic Regression'
author: "Henri V"
date: "2023-11-12"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = F)
```

# Assignment 3: Logistic Regression


## Install required packages

* Note that running this procedure requires:
  * Tidyverse
  * gt
  * here

```{r,include=F}

if (!require(here) == T) {
  install.packages("here") #Checks if the package is NOT installed. If this evaluates to TRUE, runs installation.  
}
library(here)


if (!require(tidyverse) == T) {
  install.packages("tidyverse")
}
library(tidyverse)

if (!require(gt) == T) {
  install.packages("gt")   
}
library(gt)
```

## Reading source data into R

* First off, we read the required source data into R. 
  * A separate script was created for this.
  * Data frame "alc" contains the necessary data
  
```{r,  echo=T}
source(here("Scripts/create_data_assignment3.R"))
```

## Forming initial hypotheses

* In this excercise, we try to find out if there is a connection between alcohol consumption and other variables
* The meanings of each variable are explained under "Additional variable information" [here](https://archive.ics.uci.edu/dataset/320/student+performance) 
* The following 4 are chosen and initial hypotheses for the proposed relationship formed:
  * _absences_: could high rates of alcohol consumption be affiliated with high rates of absences from school?   
  * _health_: hypothesis is that people consuming higher doses of alcohol may have a worse health condition than those who abstain?   
  * _failures_: increased likelihood of course failures among heavy drinkers? 
  * _famrel_: increased likelihood of heavy drinking leading to family conflicts (poorer family relations?) 
  
  
## Summary table for variables, graphical examinatons  

```{r, echo}  
taulukko<-alc %>% group_by(high_use) %>% summarise(`Average absence` = mean(absences),
                                            `St.Dev, absences.`=sd(absences),
                                            `Average health` = mean(health),
                                            `St.Dev., health`=sd(health),
                                            `Average failure` = mean(failures),
                                            `St.Dev., failures`=sd(failures),
                                            `Average family relations` = mean(famrel),
                                            `St.Dev., family relations` = sd(famrel))
```

```{r, results=T}
gt(taulukko) %>%
  fmt_number(decimals=2) %>% cols_align("center") %>% opt_stylize(style=1, color="green")  %>% tab_header("Summary table")

```

### Absence

* on average, absence rates seem to be higher in the high alcohol consumption group than in the low consumption group (6.4 versus 3.7). More variation in the high-consuming group.  
  * Consistent with initial assumptions.

```{r}

ggplot(alc, aes(x = high_use, y = absences)) + geom_boxplot() + ggtitle("Distribution of `absences` values by group")
```

### Health

* Regarding health, differences between groups seem small.
  * Contrary to expectation, alc consumption does not seem to differ between groups?
* Mean 3.5 versus 3.7, almost the same. SD at 1.4 in both...




```{r}
ggplot(alc, aes(x = high_use, y = health)) + geom_boxplot()+ggtitle("Distribution of `health` values by group")
```

### Failure rate
* Failure rate is on average slightly higher in the high consumption group, as was assumed, but the difference is less pronounced than I was expecting.
* 0 failures seem to be the most common situation regardless of group, any other values are marginally represented in comparison. 


```{r, results=T, echo=F, message=F}
Frek<-(alc %>% group_by(high_use, failures) %>% summarise(n = n())) 

gt(Frek) %>%  
  cols_align("center") %>% 
  opt_stylize(style=1, color="green") %>% 
  tab_header("Frequency of `failure` values by group")

```



### Family relations


```{r, echo =F}

ggplot(alc, aes(x = high_use, y = famrel)) + geom_boxplot()

```

* Family relation values appear to be on average slightly poorer in the high alcohol consumption group.
* This is the sort of finding I expected to see, but the magnitude of difference between the groups is lower than I thought it would be. . 


## Logistic regression

### Defining the GLM model, and extracting the coefficients from summary

```{r}

# find the model with glm()
m <- glm(high_use ~ failures+absences+health+famrel, data = alc, family = "binomial")

# create model summary and extract the coeffs
summary(m)

coeffs<-m %>% summary() %>% coefficients()
```
* All others except family relations are positively correlated with high alcohol use. 
* Strongest positive correlation  is Failures. Also statistically significant.
* Also absences correlates positively, shows statistical significance.
* Family relations correlates negatively, is statistically significant. Pretty much as I would expect
* Health: no statistical significance seen. 
* Next step: odds ratios 
  
### Converting coeffs to odds ratios, calculating confidence intervals
  
  
```{r, results=T, echo=T, message=F}

OddsRatio<-exp(coeffs)
ConfInt<- confint(m)

Result<-cbind(OddsRatio, ConfInt)

```




```{r, results=T, echo=T}
print(Result)

```
* These estimates are on a different scale than in the more familiar linear regression
  * There, a coefficient can be interpreted directly. When the value of the independent variable increases by 1 unit, the dependent's value changes by 1 unit as a response . Not the case here.   
  * These are _odds based_ and need to be thought of in terms of likelihood, not as a direct change in the value of dependent variable's value as independent's value changes.
  
* Odds ratios quantify the strength of the relationship between dependent and independent vars. 
* If OR = 1, the odds of being a heavy drinker or not are the same, regardless of the dependent variable.
* Here, failures has the strongest link with heavy drinking. An increase of failures by 1 increases the likelihood of being a heavy drinker by almost 2-fold (1.8)
* For absences and health, the increases in odds are less pronounced. 
* The relationship with famrel is inverted in comparison to the others: Famrel increase decreases the odd of heavy drinking by 0.25

### Predictice power



